Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-28 probabilisticProgrammingMartingales.tex
cosmetic changes to proof of antitone strict supermartingale theorem

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

@@

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
\begin{calculation}
\expect{Y_n \mid T > n}
  \step[=]{}
\frac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[=]{}
\frac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{$Y_n \geq 0$ always}
\frac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{$(Y_n)_n$ is a supermartingale}
\frac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \frac{\expect{Y_0}+1}{x \epsilon(\expect{Y_0} x^{-1})} \right \rceil$, then
\(
\mathbb P[T > B_x] \leq x.
\)
As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
  \step[=]{definition of conditional expectation for a discrete variable}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon(\frac{\expect{Y_0}}{\mathbb P[T > n]})
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x\ \epsilon(\expect{Y_0} x^{-1})
\end{calculation}
therefore, by a telescoping sum, $\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0 - B_x x \epsilon(\expect{Y_0} x^{-1})} \leq -1 < 0$, which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.

\end{proof}

>>>>

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

@@

\lo{Given a (super)martingale $(Y_n)_{n \in \omega}$ and an a.s.~finite stopping time $T$ (meaning: for some $n \in \omega$, $T \leq n$ a.s.) both adapted to the same filtration, there exist supermartingale $(Y_n')_{n \in \omega}$ and antitone function $\epsilon$ such that $(Y_n')_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.}

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore

\begin{calculation}
\expect{Y_n \mid T > n}
  \step[=]{}
\dfrac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[=]{}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{$Y_n \geq 0$ always}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{$(Y_n)_n$ is a supermartingale}
\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$, then
\(
\mathbb P[T > B_x] \leq x.
\)
As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
\lo{Change of justification of the second equality below: event, not discrete r.v. 
Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{definition of expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\[
\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
\]
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}
